Causal inference is not a statistical problem

### 2023-04-04 manuscript.qmd
tweak code

@@

d_mbias %>%
  tidy_dagitty() %>%
  ggplot(
    aes(x = x, y = y, xend = xend, yend = yend)
  ) +
  geom_dag_point() +
  geom_dag_edges() +
  geom_dag_text() +
  theme_dag() +
  coord_cartesian(clip = "off") +
  ggtitle("(4) M-bias") -> p_m_bias

d_coll %>%
  tidy_dagitty() %>%
  ggplot(
    aes(x = x, y = y, xend = xend, yend = yend)
  ) +
  geom_dag_point() +
  geom_dag_edges() +
  geom_dag_text() +
  theme_dag() +
  coord_cartesian(clip = "off") +
  ggtitle("(1) Collider") -> p_coll

d_med %>%
  tidy_dagitty() %>%
  ggplot(
    aes(x = x, y = y, xend = xend, yend = yend)
  ) +
  geom_dag_point() +
  geom_dag_edges() +
  geom_dag_text() +
  theme_dag() +
  coord_cartesian(clip = "off") + 
  ggtitle("(3) Mediator") -> p_med

d_conf %>%
  tidy_dagitty() %>%
  ggplot(
    aes(x = x, y = y, xend = xend, yend = yend)
  ) +
  geom_dag_point() +
  geom_dag_edges() +
  geom_dag_text() +
  theme_dag() +
  coord_cartesian(clip = "off") + 
  ggtitle("(2) Confounder") -> p_conf

(p_coll + p_conf) /  (p_med + p_m_bias)  
```

```{r}
n <- 100
set.seed(1331)

confounding <- tibble(
  z = rnorm(n),
  x = z + rnorm(n),
  y = 0.5 * x + z + rnorm(n)
)

set.seed(229655)

collider <- tibble(
  x = rnorm(n),
  y = x + rnorm(n),
  z = 0.45 * x + 0.77 * y + rnorm(n)
)

set.seed(207052)

mediator <- tibble(
  x = rnorm(n),
  z = x + rnorm(n),
  y = z + rnorm(n)
)

set.seed(839)

m_bias <- tibble(
  u1 = rnorm(n),
  u2 = rnorm(n),
  z = 8 * u1 + u2 + rnorm(n),
  x = u1 + rnorm(n, sd = 1),
  y =  x + u2 + rnorm(n, sd = 1)
)

d <- bind_rows(confounding, collider, mediator, m_bias) %>%
  mutate(type = rep(c("(2) Confounder", "(1) Collider", "(3) Mediator", "(4) M-Bias"), each = 100))
```

```{r}
#| fig-cap: "100 points generated using the data generating mechanisms specified (1) Collider (2) Confounder (3) Mediator (4) M-Bias. The blue line displays a linear regression fit estimating the relationship between X and Y, in each case the slope is 1. "
#| label: fig-2
ggplot(d, aes(x = x, y = y)) +
  geom_point() + 
  geom_smooth(method = "lm", formula = "y ~ x") +
  facet_wrap(~type)
```

@@

```{r}
#| label: tbl-2
#| tbl-cap: "Coefficients for the exposure under each data generating mechanism depending on the model fit as well as the correlation between $X$ and $Z$."
d %>%
  nest_by(type) %>%
  mutate(`Y ~ X` = round(coef(lm(y ~ x, data = data))[2], 2),
         `Y ~ X + Z` = round(coef(lm(y ~ x + z, data = data))[2], 2),
         `Correlation of X and Z` = round(cor(data$x, data$z), 2)) %>%
  select(-data, `Data generating mechanism` = type) %>%
  knitr::kable()

@@

```{r}
#| eval: false
#| echo: true
library(tidyverse)
# devtools::install_github("LucyMcGowan/quartet")
library(quartet)

ggplot(causal_quartet, aes(x = x, y = y)) +
  geom_point() + 
  geom_smooth(method = "lm", formula = "y ~ x") +
  facet_wrap(~dataset)

>>>>

expander <- function() {
  expand_plot(
    expand_x = expansion(c(0.01, 0.05)), 
    expand_y = expansion(c(0.2, 0.2))
  )
}

@@

d_mbias %>%
  tidy_dagitty() %>%
  ggplot(
    aes(x = x, y = y, xend = xend, yend = yend)
  ) +
  geom_dag_point() +
  geom_dag_edges() +
  geom_dag_text() +
  theme_dag() +
  coord_cartesian(clip = "off") +
  ggtitle("(4) M-bias") +
  expander() -> p_m_bias

d_coll %>%
  tidy_dagitty() %>%
  ggplot(
    aes(x = x, y = y, xend = xend, yend = yend)
  ) +
  geom_dag_point() +
  geom_dag_edges() +
  geom_dag_text() +
  theme_dag() +
  coord_cartesian(clip = "off") +
  ggtitle("(1) Collider") +
  expander()  -> p_coll

d_med %>%
  tidy_dagitty() %>%
  ggplot(
    aes(x = x, y = y, xend = xend, yend = yend)
  ) +
  geom_dag_point() +
  geom_dag_edges() +
  geom_dag_text() +
  theme_dag() +
  coord_cartesian(clip = "off") + 
  ggtitle("(3) Mediator") +
  expander()  -> p_med

d_conf %>%
  tidy_dagitty() %>%
  ggplot(
    aes(x = x, y = y, xend = xend, yend = yend)
  ) +
  geom_dag_point() +
  geom_dag_edges() +
  geom_dag_text() +
  theme_dag() +
  coord_cartesian(clip = "off") + 
  ggtitle("(2) Confounder") + 
  expander() -> p_conf

(p_coll + plot_spacer() + p_conf) /  (p_med + plot_spacer() + p_m_bias)  
```

```{r}
#| fig-cap: "100 points generated using the data generating mechanisms specified (1) Collider (2) Confounder (3) Mediator (4) M-Bias. The blue line displays a linear regression fit estimating the relationship between X and Y, in each case the slope is 1. "
#| label: fig-2
library(tidyverse)
# devtools::install_github("LucyMcGowan/quartet")
library(quartet)

ggplot(causal_quartet, aes(x = x, y = y)) +
  geom_point(alpha = 0.25) + 
  geom_smooth(
    method = "lm", 
    formula = "y ~ x", 
    linesize = 1.1, 
    color = "steelblue"
  ) +
  facet_wrap(~dataset)
```

@@

```{r}
#| label: tbl-2
#| tbl-cap: "Coefficients for the exposure under each data generating mechanism depending on the model fit as well as the correlation between $X$ and $Z$."

@@

```{r}
#| eval: false
#| echo: true
#| ref.label: ["fig-2", "tbl-2"]
```
